\textsc{\wobble: a data-driven method for precision radial velocities}

### 2018-11-19 paper/paper.tex
more discussion

\begin{abstract}
% Context
Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally informative \RVs from spectra. 
Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features, which cannot be perfectly modeled. 
Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself. 
%In both cases, the full data set of all observed spectra for a given target generally contains more complete information about the underlying 
%In both cases, using the full data set of all observed spectra for a given target will yield more precise constraints on the underlying spectra and their \RVs than a traditional analysis. %\todo{(reword this to be less controversial?)}
%Telluric absorption features in stellar spectra provide particular challenges to extreme-precision radial velocity (\EPRV) exoplanet surveys. 
%Large amplitude features are often masked in data analysis pipelines, reducing the number of observable spectral features which can be used to measure an \RV shift.
%Low-amplitude features that are hard to see even in high signal-to-noise spectra may be an important contributor to the overall \RV noise budget.
% Aims
Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a simple linear model. 
We implement this method in \wobble, an open-source \python package which uses \TF in its first non-machine-learning application to astronomical data. 
In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
% Results
We recover the canonical exoplanet 51 Pegasi b, detect the secular \RV evolution of the M dwarf \Mdwarf, and retrieve the \RM for the Hot Jupiter HD 189733b. 
We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data. 
%This method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.  
\end{abstract}

@@

An alternative approach is the simultaneous modeling of both telluric and spectral features from the data. 
As the Earth's motion around the barycenter of the solar system induces a Doppler shift considerably larger than both the motion of telluric features and the size of a single pixel on the detector, these two spectra can be disentangled.
This process is well-established in the analysis of binary star systems through the development of linear models \citep[e.g.][]{Simon1994} and in a Gaussian process framework \citep{Czekala2017}.
In these cases, both spectra are assumed to be unchanging in time, which is a reasonable approximation of a stellar spectrum but not necessarily of the telluric spectrum.
A more complicated model with time variability in the telluric spectrum may provide a more accurate fit. 
Work by \citet{Artigau2014} demonstrates that a Principle Component Analysis (\acronym{PCA}) approach is an effective way of parameterizing telluric spectral variability as a low-dimensional model derived from observational data. 
\citet{Artigau2014} use a library of telluric standard observations, which requires a significant investment of observing time to build up, but in principle such a data-driven model should be possible to derive from typical stellar observations provided enough epochs of data are available.

@@

Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is a single observation taken for the purpose \citep[e.g.][]{Butler1996}. 
Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
\todo{(explain how this differs from what we'll do)}

\todo{(talk about lack of open-source code and difficulty of modifying \RV pipelines here?)}

Here we develop a linear data-driven model to infer the telluric and stellar spectra and calculate the stellar RV at each observed epoch. 
The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
Our model requires no prior knowledge of the star or the Earth's atmosphere. 
As such, it does not yield absolute measurements of \RVs, only highly precise relative measurements between epochs.

@@

The model underpinning \wobble is designed to be flexible and easily extensible to a variety of situations. 
However, a few rigid assumptions are necessary, and we outline those here.

@@

In this work we have tested \wobble exclusively using \HARPS data, but vast amounts of data from other \EPRV instruments exist and may benefit from the techniques used here. 
We now turn our discussion to ways in which other data sets may violate the assumptions made in this work and the corresponding modifications to \wobble that would enable it to operate more effectively in these regimes. 
For simplicity, we break this discussion down into three categories: inherent limitations of \wobble; assumptions that could be eliminated; and general changes that may improve the \RV precision achieved by \wobble.

\subsection{Inherent Limitations}

limitations on number of spectra: Experiment with sub-sampling 51 Peg to fewer epochs. \wobble does have many free parameters and therefore it's not going to work well for very few epochs. Also telluric sampling is key. in some regime \wobble is just not the right tool for the job.

low-SNR limitations: what we did is a hack and it should be possible to do much better with better continuum normalization. but still you'll eventually hit a failure mode.

spectrograph resolution: will this work if lines are undersampled?

\subsection{Revisiting Assumptions}

single star: this actually isn't a true assumption, it's trivial to add an extra star, or even a planetary component, to the model.

no gas cell: should be as simple as adding an extra component (Of course, gas cells exist because calibration isn't perfect...)

reliable instrument calibration: use tellurics to modify wavelength calibration? (there are relevant citations here) Spectral variability due to LSF changing? Additional parameters to the P operators, non-convex optimization

@@

\subsection{General Improvements}

>>>>

\begin{abstract}
% Context
Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra. 
Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features, which cannot be perfectly modeled. 
Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself. 
%In both cases, the full data set of all observed spectra for a given target generally contains more complete information about the underlying 
%In both cases, using the full data set of all observed spectra for a given target will yield more precise constraints on the underlying spectra and their \RVs than a traditional analysis. %\todo{(reword this to be less controversial?)}
%Telluric absorption features in stellar spectra provide particular challenges to extreme-precision radial velocity (\EPRV) exoplanet surveys. 
%Large amplitude features are often masked in data analysis pipelines, reducing the number of observable spectral features which can be used to measure an \RV shift.
%Low-amplitude features that are hard to see even in high signal-to-noise spectra may be an important contributor to the overall \RV noise budget.
% Aims
Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a linear model (in the log flux). 
The model employs a convex objective and convex regularization to keep the optimization of the spectral components fast. 
We implement this method in \wobble, an open-source \python package which uses \TF in its first non-machine-learning application to astronomical data. 
In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
% Results
We recover the canonical exoplanet 51 Pegasi b, detect the secular \RV evolution of the M dwarf \Mdwarf, and retrieve the \RM for the Hot Jupiter HD 189733b. 
We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data. 
%This method may be of particular interest for future red-optimized spectrographs aiming to detect and characterize M dwarf host stars, where telluric features are considerable at wavelengths corresponding to the peak of the stellar spectral energy distribution.  
\end{abstract}

@@

An alternative approach is the simultaneous modeling of both telluric and spectral features from the data. 
As the Earth's motion around the barycenter of the solar system induces a Doppler shift considerably larger than both the motion of telluric features and the size of a single pixel on the detector, these two spectra can be disentangled.
This process is well-established in the analysis of binary star systems through the development of linear models \citep[e.g.][]{Simon1994} and in a Gaussian process framework \citep{Czekala2017}.
In these cases, both spectra are assumed to be unchanging in time, which is a reasonable approximation of a stellar spectrum but not necessarily of the telluric spectrum.
A more complicated model with time variability in the telluric spectrum may provide a more accurate fit. 
Work by \citet{Artigau2014} demonstrates that a Principal Component Analysis (\acronym{PCA}) approach is an effective way of parameterizing telluric spectral variability as a low-dimensional model derived from observational data. 
\citet{Artigau2014} use a library of telluric standard observations, which requires a significant investment of observing time to build up, but in principle such a data-driven model should be possible to derive from typical stellar observations if the same star is observed many times at different barycentric \RV shifts.

@@

Data analysis pipelines for absorption cell instruments have long used a stellar template that is customized to the star in question, although traditionally this template is a single observation taken for the purpose \citep[e.g.][]{Butler1996}. 
Higher quality templates can be derived iteratively from the data \citep{Sato2002, Gao2016}. 
Such data-driven templates have the benefit of being both customized to the star in question and high-\SNR without requiring observing overhead. 
%\todo{(explain how this differs from what we'll do - v similar in spirit although ours is a full forward model with stellar spectrum as a latent variable)}

%\todo{(talk about lack of open-source code and difficulty of modifying \RV pipelines here?)}

Here we develop a data-driven model to infer the telluric and stellar spectra and calculate the stellar RV at each observed epoch. 
The telluric model component may vary with time in a low-dimensional manner, which is also inferred from the data. 
Our model requires no prior knowledge of the star or the Earth's atmosphere. 
As such, it does not yield absolute measurements of \RVs, only highly precise relative measurements between epochs.

Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals. 
In this sense, ours is a true forward model.

@@

The model underpinning \wobble is designed to be flexible and easily extensible to a variety of situations. 
However, a few assumptions are made in this work to simplify the implementation, and we outline those here. 
Many of these assumptions could be eliminated with relatively straightforward modifications to the method. 
We revisit these in Section \ref{s:future}.

@@

In this work we have tested \wobble exclusively using \HARPS data, which allowed us to make relatively strict assumptions about the data quality. 
We have also made a number of restrictive assumptions about the underlying physical model, for example taking each astronomical source to be a single star. 
However, there are many other \EPRV data sets which violate some of these assumptions but could nevertheless benefit from the analysis techniques used by \wobble.

We now turn our discussion to ways in which other data sets may violate the assumptions made in this work and the corresponding modifications to \wobble that would enable it to operate more effectively in these regimes. 
For simplicity, we break this discussion down into two categories: changes to the quality or type of spectroscopic data used and changes to the model. 
We additionally outline some general changes or enhancements that could be made to potentially improve the \RV precision achieved by \wobble.

\subsection{Changes to the Data}

In general, the data used in this work are high quality in terms of \SNR, spectral resolution, and quantity. 
As demonstrated for the case of \Mdwarf, \wobble is limited in its operability when the data are at low \SNR. 
\wobble implicitly assumes Gaussian noise in $\ln$ flux, which is not strictly accurate. 
This may become a substantial issue at low \SNR, making \wobble a poor tool choice for accurate modeling of very low \SNR spectra. 
\todo{(HOGG: maybe say something about how we could mitigate this with changes to continuum handling?)}

Our method may also fail at low spectral resolution. \todo{(what will happen for undersampled lines?)}

Along with minimum requirements on the spectral data quality, \wobble has some limitations driven by the number of spectra available and the sampling of those spectra. 
Some minimum quantity of spectra are necessarily simply because the spectral templates contribute many free parameters, deriving up the data points needed for inference. 
The sampling of those spectra are also important because the stellar lines must undergo substantial \RV variations to disentangle them from the telluric features, which is only feasibly achieved by observing throughout the season to take advantage of the $\sim 30$ \kms \BERV shift. 
This requirement could be partially mitigated by modeling multiple stars with a shared telluric spectral template, which we discuss in Section \ref{s:improvements}. 
However, in general the \wobble method fundamentally requires a substantial number of high quality spectra to perform reliably.

Another key assumption made in this work that there is no gas cell represented in the data. 
While this is true for \HARPS, other instruments like \acronym{HIRES}, \acronym{APF}, and \acronym{PFS} include a gas cell set in optical path to imprint its absorption spectrum on the observed spectra \todo{(cite)}. 
Applying \wobble to spectra that include the imprint of a gas cell is technically trivial. 
The gas absorption lines are fixed features at the observatory rest frame which multiply into the spectrum. 
This makes them nearly indistinguishable from the mean telluric spectrum, and they should in fact be absorbed into this component. 
If the observations in question cover a range of different airmasses, it may be necessary to explicitly add a third model component to represent the gas cell; this component would be identical to the tellurics without the airmass scaling.

Of course, absorption cells in general exist because the instantaneous calibration of the spectrograph in question is not reliable at the required \RV precision level. 
This means that a gas cell instrument will generally violate other, more critical assumptions about the data quality.

It is likely that for such an instrument, our assumption that the wavelength solution is perfect does not hold. 
One simple way to deal with this would be to allow the observatory rest frame component(s) of the model to change in \RV. 
The stellar \RV measurement would then be differential with respect to the effective spectrograph \RV. 
\todo{(TO DO: add some citations to using tellurics for this)}

Spectral variability due to LSF changing? Additional parameters to the P operators, this would go into the non-convex optimization part (ask Hogg to throw in some words earlier on about location vs linear parameters)

\subsection{Changes to the Model}

The test data used in this work were all instances of a single, bright star, so that only one stellar component was necessary to model the astronomical source. 
Other data sets may require a more complex model, including additional stars, planetary spectra, or non-negligible sky background. 
There are some trade-offs associated with adding these components to the model, which we discuss here.
%Some or all of these additional components may also vary with time. 
%Here we briefly discuss how such modifications might be made to the \wobble model to make it applicable to a wider variety of data.

One important feature of \wobble is that it models all fluxes in the log-space. 
When the model consists solely of multiplicative components, as is the case for a single star + telluric absorption, this choice makes the optimization convex at fixed \RV and computationally tractable \todo{(HOGG: edit plz)}. 
If we chose instead to convert the $y$ data vectors to linear fluxes and use a model consisting only of additive components, the same equations and algorithms would apply. 
Example applications of such a model might be a multiple star system or a faint star with substantial sky background present in the observed spectra. 
If, however, the model needed to incorporate both additive and multiplicative components, major structural changes would be necessary and much of the optimization speed would be lost. 
This means that at present, additive components may only be easily modeled by either fixing or neglecting the multiplicative components. 
\todo{(HOGG: say things about hacks)}

Another way in which the \wobble model may fail for some data sets is in its assumptions about spectral variability.

@@

\subsection{General Changes}
\label{s:improvements}
